---
title: "R’s `data.table` package"
author: "Statistics 506, 2017"
date: ""
output: 
  html_document:
      theme: journal
      highlight: pygments
      css: styles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```

[Stats 506 Homepage](./index.html)

## About `data.table`

The `data.table` package in R provides an extension of the`data.frame` class that
aims to be both more computationally and memory efficient.  It is paricularly well suited
for large in-memory data sets and utilizes _indexed keys_ to allow quick search, subset,
and aggregrate by group operations.  


## Creating _data.table_ objects

The `data.table` package provides a function `fread` for reading delimited files
like `read.table` and `readr::read_delim`, but returns a _data.table_ object instead.
As with the _tibble_ class, _data.table_ inherits from _data.frame_.  Here is an example
using the 2014 New York City Flights data.

```{r}
library(data.table)
nyc14 = fread('https://github.com/arunsrinivasan/flights/wiki/NYCflights14/flights14.csv')
class(nyc14)
```

You can also create a _data.table_ like a _data.frame_ using:

```{r}
n = 1e3
data.table(a = 1:n, b=rnorm(n), c=sample(letters, n, replace=TRUE))
```

_Question: How can we view the *print* method for class *data.table*?_

## "Indexing"

The syntax for the `data.tables` package is inspired by the bracket (`[]`) notation
for indexing matrices and data frames. At the same time, it aims to allow many common data
operations (i.e. `dplyr` verbs) to be expressed within these brackets. 
These goals are expressed at the start of the "Introduction to data.table"
[vignettte](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html).

The basic idea is `DT[i, j, by]` where we:
  +  subset or filter rows in the `i` statement, 
  +  select, transform, or create columns (variables) in the `j` statement
  +  and group with the `by` statement.  
  
Additional operations can also be expressed within the brackets.  Rememer that,
even for a _data.frame_ or _matrix_ the left bracket `[` is actually a **function**:

```{r}
`[`
```

Below are some examples on the `nyc14` _data.table_ created using `fread()` above:

```{r}
nyc14
```


### Subsetting

Find all flights from LGA to DTW:

```{r}
lga_dtw = nyc14[origin == 'LGA' & dest == 'DTW',]
```

Get the first and last rows of `lga_dtw`:

```{r}
lga_dtw[c(1,.N)]
```

In the above, we used `.N` to index the last row.  This is a special symbol defined
by `data.table` to hold the number of rows or observations in the "current" group.
"Current" here refers to the scope in which it is used.

Also, notice the difference from standard _data.frame_ or matrix subsetting:

```{r}
# Indexing a matrix
ld_mat = as.matrix(lga_dtw)
ld_mat[c(1,nrow(ld_mat))]

# Indexing a data.fame
ld_df = as.data.frame(lga_dtw)
dim(ld_df[1:2])  # What would happen if we called: ld_df[c(1,nrow(ld_df))] ?
```

You can also use the "i" clause to order a _data.table_:

```{r}
lga_dtw[order(-month,-day,dep_time)]
```

### Column Selection

Get the departure and arrival times, flight number, and carrier for all flighs
from LGA to DTW:

```{r}
nyc14[origin == 'LGA' & dest == 'DTW', list(dep_time, arr_time, carrier, flight)]
```

Notice the use of `list()` to select columns. A synonym for `list()` within
`data.table` is  `.()` to save typing and enhance readability:

```{r}
nyc14[origin == 'LGA' & dest == 'DTW', .(dep_time, arr_time, carrier, flight)]
```


Columns can also be selected using
a character vector of column names.
```{r}
nyc14[origin == 'LGA' & dest == 'DTW', c("dep_time", "arr_time", "carrier", "flight")]
```

### Computing with Columns

The "j"-clause can be used to compute with column variables like "dplyr::summarize".
Below, we find the upper and lower quintiles of departure delays for fligths between LGA and DTW
during this period:

```{r}
lga_dtw[,.(p20=quantile(dep_delay,.2), p80=quantile(dep_delay,.8))]
```

The "j"-clause can also be used to compute with column variables much like "dplyr::transmute".
Here, we create new columns indicating whether the arrival or departure delays were
greater than 15 minutes:

```{r}
nyc14[,.(delay15 = 1*{dep_delay > 15 | arr_delay > 15})]
```

To get behvaior like "dplyr::mutate" we need [reference semantics](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-reference-semantics.html).

First, using the special function "`:=()`" we can add the `delay15` above. 

```{r}
nyc14[,`:=`(delay15 = 1L*{dep_delay > 15 | arr_delay > 15})]
nyc14
```

Pay close attention to the first line of the code above. Why didn't it print and
why didn't we need an assignment? Also, what is the role of that "L"?f

Another way to accomplish this is as below. Which do you find more intuitive?

```{r}
nyc14[,"delay30" := 1L*{dep_delay > 30 | arr_delay > 30}]
nyc14
```

### by

### `.SD`

## Copies

```{r}
DT1 = data.table(A=5:1,B=letters[5:1])
DT2 = DT1
DT3 = copy(DT1)

DT1[,C := 2*A] 
DT1
DT2 
DT3
```

## keys

## Reference Semantics

[vignette](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-reference-semantics.html)

## Resources 

  + Professor Shedden's [Large data sets in R](http://dept.stat.lsa.umich.edu/~kshedden/Courses/Stat506/r_large_files/)
  
  + The `data.table` [vignettes](https://cran.r-project.org/web/packages/data.table/) on CRAN
  
  + The project [homepage](https://github.com/Rdatatable/data.table/wiki)
   for data.table
  
  + A short introduction from Dr. Clark's [workshop](https://m-clark.github.io/data-processing-and-visualization/08_datatable.html)
  
  + See "Large Memory and out-of-memory data" on CRAN's [HPC task view](https://cran.r-project.org/web/views/HighPerformanceComputing.html)
  
  + The [`dtplyr`](https://github.com/hadley/dtplyr) package provides `dplyr` methods for
  `data.table` objects (objects of _class_ data.table). That is `dtplyr` is the
  `data.table` _backend_ for `dplyr` -- the part facing away from the interactive user.