---
title: "R’s `data.table` package"
author: "Statistics 506, 2017"
date: ""
output: 
  html_document:
      theme: journal
      highlight: pygments
      css: styles.css
      toc: true
      toc_depth: 3
      toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
library(tidyverse); library(magrittr); library(microbenchmark)
```

[Stats 506 Homepage](./index.html)

## About `data.table`

The `data.table` package in R provides an extension of the`data.frame` class that
aims to be both more computationally and memory efficient.  It is particularly well suited
for large in-memory data sets and utilizes _indexed keys_ to allow quick search, subset,
and aggregate by group operations.  


## Creating _data.table_ objects

The `data.table` package provides a function `fread` for reading delimited files
like `read.table` and `readr::read_delim`, but returns a _data.table_ object instead.
As with the _tibble_ class, _data.table_ inherits from _data.frame_.  Here is an example
using the 2014 New York City Flights data.

```{r}
library(data.table)
nyc14 = fread('https://github.com/arunsrinivasan/flights/wiki/NYCflights14/flights14.csv')
class(nyc14)
```

You can also create a _data.table_ like a _data.frame_ using:

```{r}
n = 1e3
data.table(a = 1:n, b=rnorm(n), c=sample(letters, n, replace=TRUE))
```

_Question: How can we view the *print* method for class *data.table*?_

## "Indexing"

The syntax for the `data.tables` package is inspired by the bracket (`[]`) notation
for indexing matrices and data frames. At the same time, it aims to allow many common data
operations (i.e. `dplyr` verbs) to be expressed within these brackets. 
These goals are expressed at the start of the "Introduction to data.table"
[vignettte](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html).

The basic idea is `DT[i, j, by]` where we:
  +  subset or filter rows in the `i` statement, 
  +  select, transform, or create columns (variables) in the `j` statement
  +  and group with the `by` statement.  
  
Additional operations can also be expressed within the brackets.  Remember that,
even for a _data.frame_ or _matrix_ the left bracket `[` is actually a **function**:

```{r}
`[`
```

Below are some examples on the `nyc14` _data.table_ created using `fread()` above:

```{r}
nyc14
```


### Subsetting

Find all flights from LGA to DTW:

```{r}
lga_dtw = nyc14[origin == 'LGA' & dest == 'DTW',]
```

Get the first and last rows of `lga_dtw`:

```{r}
lga_dtw[c(1,.N)]
```

In the above, we used `.N` to index the last row.  This is a special symbol defined
by `data.table` to hold the number of rows or observations in the "current" group.
"Current" here refers to the scope in which it is used.

Also, notice the difference from standard _data.frame_ or matrix sub-setting:

```{r}
# Indexing a matrix
ld_mat = as.matrix(lga_dtw)
ld_mat[c(1,nrow(ld_mat))]

# Indexing a data.fame
ld_df = as.data.frame(lga_dtw)
dim(ld_df[1:2])  # What would happen if we called: ld_df[c(1,nrow(ld_df))] ?
```

You can also use the "i" clause to order a _data.table_:

```{r}
lga_dtw[order(-month,-day,dep_time)]
```

### Column Selection

Get the departure and arrival times, flight number, and carrier for all flights
from LGA to DTW:

```{r}
nyc14[origin == 'LGA' & dest == 'DTW', list(dep_time, arr_time, carrier, flight)]
```

Notice the use of `list()` to select columns. A synonym for `list()` within
`data.table` is  `.()` to save typing and enhance readability:

```{r}
nyc14[origin == 'LGA' & dest == 'DTW', .(dep_time, arr_time, carrier, flight)]
```


Columns can also be selected using
a character vector of column names.
```{r}
nyc14[origin == 'LGA' & dest == 'DTW', c("dep_time", "arr_time", "carrier", "flight")]
```

We can deselect columns using negation (`-` or `!`) if we also set "with" to false:

```{r}
nyc14 = nyc14[, -c("tailnum"), with=FALSE]

nyc14 = nyc14[, !c("cancelled", "year", "day", "hour", "min"), with=FALSE]
nyc14
```

### Computing with Columns

The "j"-clause can be used to compute with column variables like "dplyr::summarize".
Below, we find the upper and lower quintiles of departure delays for flights between LGA and DTW
during this period:

```{r}
lga_dtw[,.(p20=quantile(dep_delay,.2), p80=quantile(dep_delay,.8))]
```

The "j"-clause can also be used to compute with column variables much like "dplyr::transmute".
Here, we create new columns indicating whether the arrival or departure delays were
greater than 15 minutes:

```{r}
nyc14[,.(delay15 = 1*{dep_delay > 15 | arr_delay > 15})]
```

To get behavior like "dplyr::mutate" we need [reference semantics](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-reference-semantics.html).

First, using the special function "`:=()`" we can add the `delay15` above. 

```{r}
nyc14[,`:=`(delay15 = 1L*{dep_delay > 15 | arr_delay > 15})]
nyc14
```

Pay close attention to the first line of the code above. Why didn't it print and
why didn't we need an assignment? Also, what is the role of that "L"?

Another way to accomplish this is as below. Which do you find more intuitive?

```{r}
nyc14[,"delay30" := 1L*{dep_delay > 30 | arr_delay > 30}]
nyc14
```

### by  

To perform operations group-wise use a `by` argument after the "j" statement.

```{r}
nyc14[,"delay30" := 1L*{dep_delay > 30 | arr_delay > 30}]
# Find percent of flights with delays of 15 minutes or more by carrier
nyc14[,.(del30_pct = 100*mean(delay30)), by=carrier]
```

We can use a list to specify multiple grouping variables.  

```{r}
# Find percent of flights with delays of 15 minutes or more by carrier and origin.
nyc14[,.(del30_pct = 100*mean(delay30)), by=.(carrier, origin)]
```

### Chaining

As with standard `data.frame` indexing,  we can compose `data.table` bracketed
expressions using chaining.

```{r}
## Find max departure delay by flight among all flights from LGA to DTW
## Then, select flights within the shortest 10% of max_delay
nyc14[origin=='LGA' & dest=='DTW', .(max_delay = max(dep_delay)), 
      by = .(carrier, flight)
     ][, .(carrier, flight, max_delay, max_delay_q10 = quantile(max_delay, .1)) 
     ][max_delay < max_delay_q10, -"max_delay_q10", with=FALSE
     ]
```

If you prefer pipes `%>%` for clarity, you can use them by appending a `.` before
the opening bracket:

```{r}
# Same as above, but using magrittr::%>%
  nyc14[origin=='LGA' & dest=='DTW', .(max_delay = max(dep_delay)), by = .(carrier, flight)] %>%
    .[, .(carrier, flight, max_delay, max_delay_q10 = quantile(max_delay, .1))] %>%
    .[max_delay < max_delay_q10, -"max_delay_q10", with=FALSE]
```


Here is a quick timing comparison between chaining, piping, and an equivalent `dplyr`
expression.  The differences are negligible for this data size.  
```{r}
# data.table chaining
f_chain <- function(){
  nyc14[origin=='LGA' & dest=='DTW', .(max_delay = max(dep_delay)), 
      by = .(carrier, flight)
     ][, .(carrier, flight, max_delay, max_delay_q10 = quantile(max_delay, .1)) 
     ][max_delay < max_delay_q10, -"max_delay_q10", with=FALSE
     ]
}

# data.table with pipe
f_pipe <- function(){
  nyc14[origin=='LGA' & dest=='DTW', .(max_delay = max(dep_delay)), by = .(carrier, flight)] %>%
    .[, .(carrier, flight, max_delay, max_delay_q10 = quantile(max_delay, .1))] %>%
    .[max_delay < max_delay_q10, -"max_delay_q10", with=FALSE]
}

# dplyr pipe
f_dplyr <- function(){
   nyc14 %>% 
    filter(origin=='LGA' & dest=='DTW') %>%
    group_by(carrier, flight) %>%
    summarize(max_delay = max(dep_delay)) %>%
    ungroup() %>%
    mutate(max_delay_q10 = quantile(max_delay, .1)) %>%
    filter(max_delay < max_delay_q10) %>%
    select(-max_delay_q10)
}

# compare timing
microbenchmark(f_chain, f_pipe, f_dplyr, times=1e3L) 
```

### `.SD`

Recall that the special symbol `.N` contains the number of rows in each 
subset defined using `by`.  

```{r}
# How many total flights by each carrier to DTW?
nyc14[dest=='DTW', .N, by=carrier]
```

There is another special symbol `.SD` which references the entire _subset of data_
for each group.  It is itself a `data.table` as we can see using the code below.

```{r}
nyc14[dest=='DTW', .(rows=nrow(.SD), cols=ncol(.SD), n=.N), by=carrier]
```

```{r}
nyc14[dest=='DTW', print(.SD), by=carrier]
```

We can pass an additional argument `.SDcols` to the bracketing function to 
limit the columns in `.SD`. Note that `.SD` does not contain the grouping columns.

```{r}
nyc14[dest=='DTW', .(rows=nrow(.SD), cols=ncol(.SD), n=.N), by=carrier,
      .SDcols = c("origin", "dest", "flight", "dep_time")]
```

```{r}
nyc14[dest=='DTW', print(.SD), by=carrier, .SDcols = c("origin", "dest", "flight", "dep_time")]
```

This can be useful in the "j" statement because it allows use `lapply` or other
functions returning a list to.

```{r}
# What is the mean departure & arrival delay for each flight to DTW?
nyc14[dest=='DTW', lapply(.SD, mean), by=.(origin, dest, carrier, flight),
      .SDcols = c("arr_delay", "dep_delay")]
```

Columns can also be specified as ranges in `.SDcols`.  
```{r}
nyc14[dest=='DTW', lapply(.SD, mean), 
      by=.(origin, dest, carrier, flight), 
      .SDcols = arr_delay:dep_delay]
```

## Copies

One of the goals of the `data.table` package is to use less memory. This is achieved
in part by preferring "shallow" copies by reference over "deep copies" by value
when appropriate. When an object is copied by _reference_ it shares physical memory
address with the object it is copied from.  This is more efficient, but _may_ lead to confusion 
as changing the value in memory also changes what is pointed to by both objects.

In the example below, we create a `data.table` `DT1` and then create a copy `DT2`.
Typical R objects would be copied by value, but `DT2` is copied by reference.  We
can ask for a copy by value explicitly using `copy()`.  
```{r}
DT1 = data.table(A=5:1, B=letters[5:1])
DT2 = DT1         # Copy by reference
DT3 = copy(DT1)   # Copy by value
DT4 = DT1[,.SD]
  
DT1[,C := 2*A]    # Create a new column 
DT1
DT2 
DT3
```

After updating `DT1` to include a new column, `C`, the column appears in `DT2` 
as well because `DT1` and `DT2` refer to the same object.  

## Reference Semantics

In the last example above we used
[reference semantics](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-reference-semantics.html) to create a new column in `DT1` without copying the other other columns and
reassigning to a new `DT1` object.  

A similar notation using `:=` as a function is below.

```{r}
DT1[,`:=`(D=C+1)]
```

One way in which this is useful is to modify subsets of a `data.table` without
re-allocating the entire thing.

```{r}
# Truncated all arrival delays at 0
nyc14[arr_delay<0, arr_delay := 0][]
range(nyc14$arr_delay)
```

We can also delete columns by reference using `NULL`:

```{r}
nyc14[,c("month") := NULL]
# i.e. nyc14$month = NULL
```

We can use this with `by` to accomplish tasks such as adding a column showing
the maximum departure delay by flight.

```{r}
nyc14[,max_dep_delay := max(dep_delay), by=.(carrier, flight)][]
```

With the functional form you can create multiple columns by reference.
```{r}
nyc14[, `:=`(max_dep_delay = max(dep_delay),
             min_dep_delay = min(dep_delay)        
            ),
      by=.(carrier, flight)
     ][]

```

The last set of empty brackets above is a short-hand for a subsequent call
to `print(nyc14)`.

## Keys and Indexing

The material on in this section is based on the vignette [here](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-keys-fast-subset.html).

Above we used "indexing" in a generic sense to mean "subsetting".  What we mean by "indexing" here
is more specific and [technical](https://en.wikipedia.org/wiki/Database_index): we create an
_indexed_ data table by designating specific columns as _keys_ and sorting the table by these keys to
create more efficient lookups and aggregations.  

We can designate one or more columns as a _key_ in `data.table` using `setkey()`.

```{r}
setkey(nyc14, origin) #also, setkeyv(nyc14, "origin") if character is preferred.
key(nyc14)
```

In `data.table` when we designate columns as keys, the rows are re-ordered by _reference_ in
_increasing_ order. In other words, the original columns are left _as is_ in memory but a new
column `key` is created to store the order of the `data.table`. This works something like this:

```{r}
# Example of keys on a data-frame
df = data.frame(ID = sample(LETTERS,100,replace=TRUE), value=rpois(100, 10))
df$key = order(df$ID)
head(df)
head(df[df$key, ])
```

After a key has been set, we can compute in the "i"-statement using lists:

```{r}
nyc14[.("LGA")]
```

We can have more than one column contribute to the order used to form the key.

```{r}
# key by origin and destination
setkey(nyc14,origin,dest)
key(nyc14)

# Get flighs of any origin to DTW
nyc14[.(unique(origin),'DTW')]
```

We can of course combine this with 'j' and 'by' statements.

```{r}
# Find the median departure delay for all flights to DTW
nyc14[.(unique(origin),'DTW'), 
      .(med_dep_delay = as.numeric(median(dep_delay)), n=.N),
      by=.(origin, dest, flight)
     ] %>%
  .[order(origin, med_dep_delay, -n)] %>% # Order
  .[n>42]  # Limit to at least ~1 flight per week
```

What is the advantage of indexing by setting a key? 

```{r key_example}
# Generate two columns of data with 10 million rows
N = 1e8
DT = data.table(group = sample(1:26, N, replace=TRUE))
DT = DT[ ,.(count=rpois(.N,group)), keyby=group]

# See size of this data
print(object.size(DT),units='MB')

## Unkeyd approach
tm0 = system.time({
  ans0 <- DT[group==1 | group==26, .(lambda_hat = mean(count)), by=group]
})

## Alternate unkeyed approach
tm1 = system.time({
  ans1 <- DT[, .(lambda_hat = mean(count)), by=group][group==1 | group==26]
})

# Set key
setkey(DT,group)

# keyed approach
tm2 = system.time({
  ans2 <- DT[.(c(1,26)),.(lambda_hat = mean(count)), by=group]
})

# Compare timings (seconds)
rbind(tm0,tm1,tm2)[,'elapsed']

rm(DT,tm0,tm1,tm2)
```

## Reshaping data tables

We did not cover this in class, but you may wish to read this [vignette](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-reshape.html).


## Resources 

  + Professor Shedden's [Large data sets in R](http://dept.stat.lsa.umich.edu/~kshedden/Courses/Stat506/r_large_files/)
  
  + The `data.table` [vignettes](https://cran.r-project.org/web/packages/data.table/) on CRAN
  
  + The project [homepage](https://github.com/Rdatatable/data.table/wiki)
   for data.table
  
  + A short introduction from Dr. Clark's [workshop](https://m-clark.github.io/data-processing-and-visualization/08_datatable.html)
  
  + See "Large Memory and out-of-memory data" on CRAN's [HPC task view](https://cran.r-project.org/web/views/HighPerformanceComputing.html)
  
  + The [`dtplyr`](https://github.com/hadley/dtplyr) package provides `dplyr` methods for
  `data.table` objects (objects of _class_ data.table). That is `dtplyr` is the
  `data.table` _backend_ for `dplyr` -- the part facing away from the interactive user.
  
  + [Practice Exam Questions](./DplyrVsDT.pdf). A similar question will appear on the final exam. Here are example [solutions](./DplyrVsDT_solutions.html).
  