---
title: "Home Types Case Study"
author: "Statistics 506, Fall 2017"
date: ""
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```

## Case Study

In this document we will use the RECS data as a case study for introducing
"split, apply, combine" and several R packages from the [tidyverse](https://www.tidyverse.org/).

We will take a brief look at `readr` and then illustrate some of the functionality 
of `dplyr` and `tidyr`. You should read the following introduction to [dplyr](https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html) and work through the examples there on your own. 

# {.tabset}
## readr

The readr package is one of several alternatives to `read.csv()` and `read.table` in the base R package `utils`. Let's do a quick comparison of the two
using `system.time()`.

```{r, echo=TRUE,warning=FALSE}
## load (install) readr
#install.packages(readr)
library(readr)

# setwd('~/Stats506/')

## Base R read.csv is a wrapper around read.table
system.time({
  recs_df = read.table('./recs2009_public.csv',
                       stringsAsFactors = FALSE, header=TRUE, 
                       sep=',')
}
)
class(recs_df)

## readr 
system.time({
  suppressMessages({
  recs_tib = read_delim('./recs2009_public.csv', delim=',',
                        col_names=TRUE)
  })
})
class(recs_tib)
```

In the example above, note the use of braces `{}` to encapsulate an 
expression with assignment within the call to `system.time()`. 


## Split, Apply, Combine and dplyr

One of the most frequently used patterns in data analysis and management is performing some operation repeatedly over subsets of
data. This is often called "split-apply-combine" and is an approach emphasized by the R package 
[dplyr](https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html). You should read the introduction to [dplyr](https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html) and work through the examples there. 

### Case Study: Home types by state

We will use the RECS data imported above to demonstrate some of the functionality of 
dplyr.  Let's set out to answer the question, which state has the highest proportion
of single-family attached homes?

Let's begin by creating a clean data frame with the necessary variables. We'll
need to keep the two variables of interest and the sample weight.

```{r, subset}
## Subset recs to home-type, state, and sample weight.
recs_homes = transmute(recs_tib, State=REPORTABLE_DOMAIN, Type=TYPEHUQ, 
                       Weight = NWEIGHT)
recs_homes
```

Next, let's clean up the values to something more easily interpreted. We will write
functions for the mappings so we can reuse them later. The functions will begin
with the `decode_` prefix so we can remember them as a group.

```{r decode_states}
# Functions to decode states or reportable_domains from RECS.

# decode_state returns a single value
decode_state = function(x){
  if(!is.numeric(x)) stop('decode_states expects numeric input indexed from 1!')
  
  switch(x,
      "CT, ME, NH, RI, VT", "MA", "NY", "NJ", "PA", "IL", "IN, OH", "MI", "WI",
      "IA, MN, ND, SD", "KS, NE", "MO", "VA", "DE, DC, MD, WV", "GA",
      "NC, SC" , "FL", "AL, KY, MS", "TN", "AR, LA, OK",
      "TX", "CO", "ID, MT, UT, WY", "AZ", "NV, NM",
      "CA", "AK, HI, OR, WA"
    )
}

# this is a wrapper to an apply call
decode_all_states = function(x){
  sapply(x, decode_state)
}
```


```{r decode_type}
# Functions to decode housing type
decode_house_type = function(x){
  if(!is.numeric(x)) stop('decode_house_type expects numeric input indexed from 1!')
  
  switch(x,
         'MobileHome',
         'SingleFamilyDetached',
         'SingleFamilyAttached',
         'ApartmentFew',
         'ApartmentMany'
         )
}

decode_all_house_types = function(x){
  sapply(x,decode_house_type)
}
```

Next we `mutate` our data frame, replacing our numeric codes with decoded values.
```{r mutate}
recs_homes = mutate(recs_homes, 
                    State=decode_all_states(State),
                    Type=decode_all_house_types(Type))
recs_homes
```

#### group_by and summarize
Recall that we are interested in computing the proportion of each housing type
by state. We can do this using the split, apply, combine, paradigm -- in dplyr
and other languages this is accomplished using a group operation with an aggregation
function.  In dplyr we use `group_by` to group and `summarize` to aggregate. 

First the grouping,
```{r group}
recs_homes_group_states = group_by(recs_homes, State, Type)
recs_homes_group_states
```

and now the aggregation 
```{r summarize}
recs_type_state_sum = summarize(recs_homes_group_states, Homes=sum(Weight))
recs_type_state_sum
```

Pay close attention to the change in grouping.  When `summarize` is called we
lose the most nested group.

#### spread, select, arrange

To proceed, let's reshape the data to have one row per state.  We can do this
using the `tidyr::spread` function.

```{r}
# Reshape to wide format for easy computation of proportions
recs_type_state = tidyr::spread(recs_type_state_sum,Type,Homes)

# Compute proportions
recs_type_state = mutate(recs_type_state,
                         Total = ApartmentFew + ApartmentMany + MobileHome +
                           SingleFamilyAttached + SingleFamilyDetached,
                         
                         ApartmentFew         = 100*ApartmentFew/Total,
                         ApartmentMany        = 100*ApartmentMany/Total,
                         MobileHome           = 100*MobileHome/Total,
                         SingleFamilyAttached = 100*SingleFamilyAttached/Total,
                         SingleFamilyDetached = 100*SingleFamilyDetached/Total
                         )

# Drop total
recs_type_state = select(recs_type_state,-Total)

# Sort by Single Family Attached
recs_type_state = arrange(recs_type_state, SingleFamilyAttached)

# Use desc() to sort in descending order
recs_type_state = arrange(recs_type_state, desc(SingleFamilyAttached))
recs_type_state
```

A nicely formatted table would be rounded to a whole or tenth of a percent. In
general, it is better to round only for display and not in the data itself to prevent
rounding error from accumulating.

```{r}
knitr::kable(recs_type_state,digits=1,caption='Proprtion of home types by State(s).')
```

#### Filter 
Next we take a quick look at just Michigan to demonstrate the use of `filter`.

```{r}
recs_type_state %>% filter(State=='MI')
```

We might also want to find all states with at least 25% of people living in apartments,

```{r}
recs_type_state %>% filter( {ApartmentFew+ApartmentMany} >= 25)
```

###  Caution

If you look at the bottom of the home-types by state table, you'll notice Illinois
has `NA` rather than values.  What happened? 

Looking at the state/type summaries we will notice that there is no mobile home
type for Illinois.  
```{r}
recs_type_state_sum %>% filter(State=='IL')
```

In our previous call to mutate, we didn't guard against this.  We can do something
safer to get more robust results. One way to do this would be to use the `rowwise`
function which ensures vectorized R functions such as `sum` are interpreted to act row by row.
This illustrated under the "pipes" tab.


## Pipes `%>%`

At then end of the case study, you may have notice an odd symbol `%>%`. This "pipe"
operator can be used to pass a data frame implicitly to dplyr functions.  The utility
of this lies in allowing us to string dplyr functions together in such a way that
they read in the same order as performed rather from the inside out.  This also
prevents us from needing to keep track of so many intermediate objects.  Together
these two properites make our code cleaner -- making it easier to understand and 
debug.

Here is the entire example as a single piped chain,

```{r}
home_type_prop = recs_tib %>% 
  transmute(State=REPORTABLE_DOMAIN, Type=TYPEHUQ, Weight = NWEIGHT) %>%  
  mutate(State=decode_all_states(State), Type=decode_all_house_types(Type)) %>%
  group_by(State, Type) %>%
  summarize(Homes=sum(Weight)) %>%
  tidyr::spread(Type,Homes) %>%
  mutate(Total = ApartmentFew + ApartmentMany + MobileHome + 
                 SingleFamilyAttached + SingleFamilyDetached,
         ApartmentFew         = 100*ApartmentFew/Total,
         ApartmentMany        = 100*ApartmentMany/Total,
         MobileHome           = 100*MobileHome/Total,
         SingleFamilyAttached = 100*SingleFamilyAttached/Total,
         SingleFamilyDetached = 100*SingleFamilyDetached/Total
        ) %>%
  select(-Total) %>%
  arrange(desc(SingleFamilyAttached))

# Change options for display
options(list('digits'=2,dplyr.width=Inf))
home_type_prop
```

Here is the same example using `rowwise` to avoid `NA` from missing values in 
our sum (assuming that is the behavior we want.)

```{r }
home_type_prop_rowwise = recs_tib %>% 
  transmute(State=REPORTABLE_DOMAIN, Type=TYPEHUQ, Weight = NWEIGHT) %>%  
  mutate(State=decode_all_states(State), Type=decode_all_house_types(Type)) %>%
  group_by(State, Type) %>%
  summarize(Homes=sum(Weight)) %>%
  tidyr::spread(Type,Homes) %>%
  rowwise() %>%  # to guard against NA when summing "Total"
  mutate(Total = sum(ApartmentFew, ApartmentMany, MobileHome,
                 SingleFamilyAttached, SingleFamilyDetached, na.rm=TRUE),
         ApartmentFew         = 100*ApartmentFew/Total,
         ApartmentMany        = 100*ApartmentMany/Total,
         MobileHome           = 100*MobileHome/Total,
         SingleFamilyAttached = 100*SingleFamilyAttached/Total,
         SingleFamilyDetached = 100*SingleFamilyDetached/Total
        ) %>%
  select(-Total) %>%
  arrange(desc(SingleFamilyAttached))

## In an R script, options() sets global options.  
## These seem to be wiped between code chunks in R-Markdown.
getOption('digits')
options(list('digits'=2,dplyr.width=Inf))
getOption('digits')

home_type_prop_rowwise %>% filter(State=="IL")
```


