---
title: "Dplyr vs Data.Table"
author: "Statistics 506, Fall 2017"
date: ""
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
library(tidyverse)
library(data.table)
nyc14 = fread('https://github.com/arunsrinivasan/flights/wiki/NYCflights14/flights14.csv')
```

## Instructions

Consider the 2014 flights data used for the `data.table` examples. For each
code snippent first provide a single-setence summary of the value(s) being computed.
Make your answer as concise and descriptive as possible. 

## Questions 

  1.  Consider the following `data.table` expression:

```{r}
tab1 =
  nyc14[ , .(n = .N), by=.(origin, dest)] %>%
    .[, .(origin, n = n, pct = n / sum(n)), by=dest] %>%
    .[pct > .75] %>%
    .[order(-pct, dest)]
```

a. Provide a one-sentence summary of what is being computed.

_This code finds all destination cities for which more than 75% of
the flights are from a single origin airport._
    
b. Provide a translation using `dplyr` syntax.

```{r}
nyc14 %>%
  group_by(dest, origin) %>%
  summarize(n=n()) %>%
  mutate(pct = n/sum(n)) %>%
  filter(pct > .75) %>%
  arrange(-pct, dest)
```

***

  2. Consider the `dplyr` code snippet below.  
  
```{r}
tab2 =
  nyc14 %>%
    group_by(origin, dest, carrier) %>%
    summarize(n = n()) %>%
    filter(n >= 80) %>%
    group_by(origin, carrier) %>%
    summarize(n = n()) %>%
    arrange(origin,-n)
```

 a. Provide a one-sentence summary of what is being computed.
 _For each carrier and origin airport, this code counts the number of destination
 cities with at least 80 observed flights._
 
 b. Provide a translation using `data.table` syntax.
 
```{r}
nyc14[, .(n=.N), by = .(origin, dest, carrier)
     ][n>=80
     ][, .(n=.N), by = .(origin, carrier)
     ][order(origin,-n)
     ]
```
 
    
***

  3.  Consider the R code snippet below.
  
```{r}
nyc14_df = as.data.frame(nyc14)
tab3 =
 with(
  with(nyc14_df,nyc14_df[grepl('HOU',dest),]),
      {
       keys = paste(carrier,month,sep=':')
       u = unique(keys)
       n = sapply(u, function(key) sum(key==keys))

       tmp = strsplit(u,':')
       carrier = sapply(tmp,function(x) x[1])
       month  = sapply(tmp,function(x) x[2])
       
       cr  = unique(carrier)
       ind = sapply(cr, function(x) grep(x,carrier, fixed=TRUE))

       data.frame(carrier,month,n)[ind,]
      }
 )
```

  a. Provide a one-sentence summary of what is being computed.
  
  _This code computes the number of flights from NYC to HOU each month
  by carrier._
    
  b. Provide a translation using `data.table` syntax.

```{r}
setkey(nyc14, dest)
nyc14['HOU', .(n = .N), by=.(carrier, month)][order(carrier,month)]
```

    
  c. Provide a translation using `dplyr` syntax.
  
```{r}
tab3c = nyc14 %>%
  filter(dest=='HOU') %>%
  group_by(carrier, month) %>%
  summarize(n = n())
```


