---
title: "Object Oriented Programming in R"
author: "Statistics 506, Fall 2017"
date: ""
output: 
    html_document:
    theme: journal
    highlight: pygments
    css: styles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Object Oriented Programming

When we attach values to names in an R environment we generally refer to 
the name and value collectively as an 'object'. [Object oriented programming](https://en.wikipedia.org/wiki/Object-oriented_programming)
is a programming paradigm built around the notions of classes, methods, and,
of course, objects. There are a wide variety of object oriented languages and R
has (at least) three _object oriented_ (OO) systems you should be aware of. 

Before digging into R's OO systems it will be helpful to define a few terms.

  + An object's __class__ defines its structure and behavior using _attributes_
   as well as its relationship with other classes. 
   
  + __Methods__ are functions that behave differently depending on the class of an
  object.
  
  + Classes are often organized into hierarchies with 
  "child" classes defined more strictly than its "parents".  Child classes often
  __inherit__ from their parents meaning a parent's structure or methods serve
  as a default when not explicilty defined for the child.
  
## The S3 system in R

The S3 system in R is based on the idea of generic functions.  The basic idea is
that a generic function is used to dispatch a class-specific method for a given
object.  Some common S3 generic functions in R inlcude, `print`, `summary`, `plot`,
`mean`, `head`, `tail`, and `str`.  If we look at the definitions for these functions, we see they are
all quite simply defined in terms of a call to `UseMethod`.

```{r}
print
summary
head
```

When `UseMethod` is called R searches for an S3 method based on the name of the
generic function and the class of its first argument.  As an example, let's construct
a matrix object `mat` and examine a call to `head(mat)`:

```{r}
mat = matrix(1:45, nrow=9, ncol=5)
class(mat)
head(mat)
```

The object `mat` has class `matrix` so `UseMethod("head")` searches for a function
(method) called `head.matrix()` to apply to `mat`:  

```{r}
head.matrix
```

You can see all the methods associated with a generic function using `methods()`. 

```{r}
methods(head)
```

When an object has more than one class, R searches successively until a suitable method is found.
```{r}
class(mat) = c('green', class(mat))
class(mat)
head(mat)
```

If a suitable method is not found, S3 generics revert to a default method if
defined and throw an error if not.  We can call some methods explicitly:

```{r}
head.matrix(mat)
```

but others such as `head.default` are not exposed.  You can view the source
code for unexposed S3 generics using `getS3method('generic','class')`.

```{r}
getS3method('head', 'default')
```

Defining a new S3 method is as simple as defining a function and naming it 
accordingly. Here we define a method `head.green`.

```{r}
head.green = function(obj) {
  # Make sure obj is an object
  if(!is.object(obj)) warning("Object 'obj' is not an object!")

  # Check if its green
  if ('green' %in% class(obj)) {
    if (length(class(obj)) > 1) {
      next_class = class(obj)[-grep('green', class(obj))][1]
      cat('This is a green ',next_class,'.\n',sep='')
    } else{
      cat('This a generic green object.\n')
    }
  } else{
    cat('The object is not green!\n')
  }
  
}
```

Now we can test it under various conditions.
```{r}
## We previously assigned
class(mat)
head(mat)

## Test head.green for generic class
class(mat) = 'green'
head(mat)

## Test on a non-green object
red_obj = 1:100
class(red_obj) = 'red'
head.green(red_obj)
head(red_obj)
```

We can also define our own S3 generic functions via `UseMethod`.

```{r}
# Generic Color finder
getColor = function(obj) {
  UseMethod("getColor")
}

# Default method 
getColor.default = function(obj) {
  # Are any classes colors?
  ind = class(obj) %in% colors()
  if (any(ind)) {
     # Yes. Return first color.
     class(obj)[which(ind)[1]]
  } else {
    # No return a random color.
    sample(colors(),1)
  }
}

# Specific method for green
getColor.green = function(obj) {
  "darkgreen"
}
```

As a quick example of how we might use this, we could define
a `col_boxplot` function to pick colors according to the class of the object passed.

```{r}
# A box plot function that uses the class attribute to define colors.
col_boxplot = function(dat, ...) {
  if(is.atomic(dat)){
    boxplot(dat, col=getColor(dat), ...)
  } else{
    col = sapply(dat, getColor)
    boxplot(dat, col=col, ...)
  }
}
```

```{r}
# Define some iid data
x = rnorm(100, 1, 1); class(x) = 'green'
y = rnorm(100, 0, 2); class(y) = 'red'
z = rnorm(100, 0, 1)
col_boxplot(list(x=x, y=y, z=z), las=1)
col_boxplot(list(z=x, y=y, z=z), las=1)
```

You should be aware that the class of the object returned by
generic functions can depend on the input class.

```{r}
class(x + y)
class(y + x)
class(mean(x))
```

### Defining an S3 class
The majority of S3 objects are simply lists plus a class attribute. As an example,
consider the class `lm` returned by the `lm` function for linear regression modeling,

```{r}
# How does head.function work?
getS3method('head','function')

# Borrow the structure to read lm line by line
lines = deparse(lm)

# Print lines involving 'list' keyword
noquote(lines[grep('list', lines)])

# Print lines involving 'class' keyword
noquote(lines[grep('class', lines)])
```


## The S4 System
The S3 system described above is very flexible making it easy to work with, but 
at the expense of the safety and uniformity of a more formal OO system.  

The S4 system is a more formal OO system in R. One key difference is that S4
classes have formal definitions and classes, methods, and generics must all be
explicitly defined as such. 

### Defining an S4 class

S4 classes are defined using the `setClass` function:
```{r}
setClass("color_vector",
   representation(
     name='character',
     data='numeric',
     color='character'
   )
)
```

Create a new instance of an S4 class using `new`:
```{r}
x = new("color_vector", name="x", color="darkgreen")
x
```

The function `new` is used above as a _constructor_ for creating an object
with the desired class.  Most S4 classes defined in packages you download
have their own constructors which you should use when defined.  We can
create a constructor by assigning the output of `setClass` a name:

```{r}
color_vector = 
 setClass("color_vector",
   representation(
     name='character',
     data='numeric',
     color='character'
   )
 )
y = color_vector(name="y", data = rnorm(100, 0, 2), color="red")
class(y)
```

### Accessing slots in an S4 object

You can access and set attributes for an S4 object using an `@` symbol, the
`slot` function, or an `attr(obj, 'name')` construction:

```{r}
## Access slots using @
x@color

# Assign some data to the data slot
x@data = rnorm(10, 1, 1)

# Check the color
slot(x, 'color')

# Change the name of x
attr(x, 'name') = 'Green Values'
names(attributes(x))
```

### S4 Methods

We can control how an object of class `color_vector` gets printed by defining 
a `show` method (the S4 equivalent of `print`).


```{r}
## This is an S4 generic
show(x)
```

```{r}
## Change how color_vector objects are shown.
setMethod('show', 'color_vector',
  function(object) {
    msg = sprintf('name: %s, color: %s\n\n', object@name, object@color) 
    cat(msg)
    cat('Data:')
    str(object@data)
    cat('\n')
  }
)
```

Now, when we call `show` on an object of class `color_vector` R will use the custom
method.

```{r}
slot(x, 'data') = rnorm(100, 1, 1)
show(x)
```

To create new S4 generics, we use `setGeneric` and  `standardGeneric`.

```{r}

```

## An S4 Example

```{r}
library(stats4)

# Generate some data
y = rpois(30, 14)

# Write a function to compute negative log likelihood
neg_log_lik <- function(lambda) - sum(dpois(y, lambda, log=TRUE))

fit = mle(neg_log_lik, start=list(lambda=mean(y)), nobs=length(y))
isS4(fit)

fit_profile = profile(fit)
class(fit_profile)
plot(fit_profile)

names(attributes(fit_profile))
diff(fit_profile@profile$lambda[,2])
```

To view the definition of an S4 method, use `getMethod`. 
```{r}
head(getMethod(profile,'mle'))
```



## Resources

  + "Object Oriented Programming" (chapter 9) in Norman Matloff's 
  _The Art of R Programming_.  

  + ["OO Field Guide"](http://adv-r.had.co.nz/OO-essentials.html) (chapter 7) in Hadley Wickham's _Advanced R_.
  
  + [The R Language](http://dept.stat.lsa.umich.edu/~kshedden/Courses/Stat506/r_language/)
  from Professor Shedden's 2016 course notes.
  
  + _Programmer's Niche: A simple class, in S3 and S4_ by Thomas Lumley on page 33 at this [link](https://cran.r-project.org/doc/Rnews/Rnews_2004-1.pdf)
  
  + The [R6 package](https://cran.r-project.org/web/packages/R6/vignettes/Introduction.html)
  provides `C++` style classes in R. 
