---
title: "Using C/C++ in R"
author: "Statistics 506, Fall 2017"
date: ""
output: 
  html_document:      
    theme: journal
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Rcpp)
```

## Resources
 
 + R's C interface
    + Chapter 15 of _The Art of R Programming_
    + Chapter 20 of _Advanced R_
    + "System and foreign language interfaces", [Section 5](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#System-and-foreign-language-interfaces)
  of _Writing R Extensions_

 + Learning C
    - [C Tutorial](https://www.tutorialspoint.com/cprogramming/)
  
 + Rcpp
    - [rcpp.org](http://rcpp.org)
    - [Gallery](http://gallery.rcpp.org)
    - [Chapter 19](http://adv-r.had.co.nz/Rcpp.html) of _Advanced R_
    - [Rcpp Quick Reference](http://dirk.eddelbuettel.com/code/rcpp/Rcpp-quickref.pdf)
    
 + Learning C++
     - [Tutorial](http://www.cplusplus.com/doc/tutorial/) from cplusplus.com
     - [Tutorial](https://www.tutorialspoint.com/cplusplus/) from tutorialspoint.com

## C and C++ Basics

  These are *compiled* languages and hence a compiler is needed to transform
 a human interpret able program into a machine executable one.  The most commonly
 use compiler is [gcc](https://gcc.gnu.org/).  This is already installed on most
 Linux distributions, including all the servers we have commonly used in this course.
 If you wish to compile C/C++ programs on your personal computer, you will need 
 to install a compiler.  
 Mac users should download [Xcode](https://developer.apple.com/xcode/) from Apple's app store.
 Windows users should investigate [Rtools](https://cran.r-project.org/bin/windows/Rtools/) or [Cygwin](https://www.cygwin.com/).

There are three important differences from R 

  + Variables and functions need to be declared before use.
  + Arrays are indexed from zero.
  + Arrays are stored in *row-major* order.
  + Statements are delimited using semicolons ";".
  + Loops have minimal overhead and hence generally do *not* need to be avoided.

### C Basics

We will review the following topics from [C Tutorial](https://www.tutorialspoint.com/cprogramming/):

  + Program Structure
  + Basic Syntax
  + Data types (Integer and Floating Point Types)
  + Variables 
  + Operators (Arithmetic, Relational, Logical and Assignment)
  + Loops
  + Arrays
  + Pointers
  + Type Casting

We will not review these as they work very much like in R:

  + Decision Making

The remaining sections are mostly beyond the scope of what we will discuss here.

## Calling C functions from R

We can call a function written in C from R using `.C()` or `.call()`.  We begin
with `.C` as it is simpler.  

### Using `.C()`

The basic recipe for using `.C` is as follows:
  
  + Write a function in `.C` that returns `void` and stores the desired result
  in one or more arguments expressly created for this purpose. Your function
  should have a `C_` prefix.
  
  + Compile your function into a shared library accessible to R using this command:
  `R CMD SHLIB my_func.c` with `my_func.c` replaced by your c program.
  
  + Within an active R session, link to the shared library using `dyn.load("my_func.so")`.
  
  + Write an R wrapper to call your C function using `.C` with the following syntax:
    `.C("C_my_func", arg1, arg2)`.  The reason to use an R wrapper is to ensure that
    the arguments passed are of the correct size and type in order to avoid potentially
    fatal errors.

 The middle two steps can be avoided by including these functions into an R
 package.  We will discuss this briefly when turning to C++ and Rcpp.

 A conceptual key to understanding how `.C` works is realizing that function arguments
 are passed as *pointers* to objects in memory rather than by value.  
    
 We will look at some examples of C coding available [here](./C_ex.tgz).
 
### Using `.Call()`

We can use `.Call()` to create and modify R level objects
directly using functions in C.  R level objects are of type `SEXP` for 
"S-expression". These types are defined in the header file `Rinternals.h` which
should be included in all C functions to be called with `.Call()`.

When we create R level objects, we must protect them from garbage collection
using `PROTECT` or a similar function.  We later `UNPROTECT` them to allow
the memory to be reallocated. 

Recall that scalars in R are length one vectors. Consequently, the must be
coerced to C type scalars to be uses as such.  This can be done using the
C functions `as*`: `asLogical`, `asInteger`, `asReal`, `CHAR(asChar())`. 

Here is a simple example using the `inline::cfunction`.
```{r seqn}
library(inline)
seqn = cfunction(signature(n_="integer", start_="integer"), 
 body="
 int i, n = asInteger(n_), start=asInteger(start_);
 SEXP out = PROTECT(allocVector(INTSXP, n));
 for(i=0; i<n; ++i){
   INTEGER(out)[i]=i+start;
 }
 UNPROTECT(1);
 return out;
")
seqn(8, 3)
```

```{r fib}
fib = cfunction(c(n_="integer"), body=
  "
  /* Declaratations */
  int n = asInteger(n_);
  SEXP out = PROTECT(allocVector(INTSXP, n));

  /* Function Body */
  INTEGER(out)[0] = 1;
  if(n > 1)  INTEGER(out)[1]=1;
  if(n > 2)
    {
      for(int i=2; i<n; i++)
        {
        INTEGER(out)[i] = INTEGER(out)[i-2] + INTEGER(out)[i-1];
        }
    }

  /* Clean up and return */
  UNPROTECT(1);
  return out;
  ")
fib(10)
```

### Modifying inputs
Be careful about modifying input arguments without duplicating first as other
R objects may point to the same location in memory owing to "copy-on-modify" semantics.  
Here is a quick example from p. 441 of _Advanced R_ using the "inline" package.

```{r add_three}
add_three = cfunction(c(x="numeric"),
  "
   REAL(x)[0] = REAL(x)[0] + 3;
   return x; 
  ")
y <- x <- 1
add_three(x)
x
y
```

Here is a corrected version in which the argument is duplicated rather
than modified in place. 
```{r add3}
add_3 = cfunction(c(x_="numeric"),
  "
   SEXP x = PROTECT(duplicate(x_));
   REAL(x)[0] = REAL(x)[0] + 3;
   UNPROTECT(1);
   return x; 
  ")
y <- x <- 1
add_3(x)
x
y
```

## Interfacing with C++ via Rcpp

The  [Rcpp](http://rcpp.org) package greatly simplifies the process of
exposing functions written in C++ to R.  

### Using sourceCpp {.tabset}

We can use the function `sourceCpp` to read a function written in C++ into
R interactively.  The function takes care of the compilation using `R CMD SHLIB` 
and automatically generates an R wrapper for the underlying function.  The shared
library and other files will be written to the directory specified by `cacheDir`
which defaults to a temporary directory for automated clean up.

To use a C++ function via sourceCpp:

  1. Write a your C++ function in a file with extension `.cpp`.
  2. In the source file, be sure to `#include <Rcpp.h>`.
  3. Designate functions exposed to R using the tag `// [[Rcpp::export]]`.
  4. Compile and source your function using `sourceCpp()` and a link to the file.

Here are some minimal examples based on generating Fibonacci numbers. You
can download the source code [here]().

#### FibCpp0

In this first example, we use the STL ("standard template library") for
vector classes.  Notice the use of `::` to specify the namespace just like
with R packages. Also pay attention to how we can handle errors using `throw`
and an error class from the STL.

```cpp
#include <Rcpp.h>

// [[Rcpp::export]]
std::vector<int> FibCpp0(int n)
{
  // Error checking
  if(n <= 0)
    {
      throw std::range_error("n must be a positive integer");
    }

  // Allocate memory
  std::vector<int> out(n);
  out[0]=1;

  // Compute additional terms
  if(n > 0)
    {
      out[1]=1;
      int i;
      for(i=2; i<n; i++)
        {
          out[i] = out[i-1] + out[i-2];
        }
    }

  return out;
}
```

```{r,error=TRUE}
#library(Rcpp)
sourceCpp('~/Stats506/FibCPP0.cpp')
tail(FibCpp0(30))
tryCatch(FibCpp0(-1),outFile=stdout())
```

#### FibCpp1
This is nearly identicaly to `FibCpp0` except we use the Rcpp class
`IntegerVector` for vectors.  

```cpp
#include <Rcpp.h>

// [[Rcpp::export]]
Rcpp::IntegerVector FibCpp1(int n)
{
  // Error checking
  if(n <= 0)
    {
      throw std::range_error("n must be a positive integer");
    }

  // Allocate memory
  Rcpp::IntegerVector out(n);
  out[0]=1;

  // Compute additional terms
  if(n > 0)
    {
      out[1]=1;
      int i;
      for(i=2; i<n; i++)
        {
          out[i] = out[i-1] + out[i-2];
        }
    }

  return out;
}
```

```{r}
sourceCpp('~/Stats506/FibCPP1.cpp')
FibCpp1(12)
```

#### FibCpp2
The final Fibonacci example here makes two changes from the previous example:

  1. We add the line `using namespace rcpp;` to avoid having to type the prefix
  `Rcpp::` when referring to the Rcpp namespace.  This is much like using 
  `library()` within R to add a package to the search path.
 
  2. We show how to print to the R console from within the C++ function using 
  `Rcout`.  This can be useful for manual debugging.

```cpp
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
IntegerVector FibCpp2(int n)
{
  // Error checking
  if(n <= 0)
    {
      throw std::range_error("n must be a positive integer");
    }

  // Allocate memory
  IntegerVector out(n);
  out[0]=1;

  Rcout << "Starting computations ... ";
  // Compute additional terms
  if(n > 0)
    {
      out[1]=1;
      int i;
      for(i=2; i<n; i++)
        {
          out[i] = out[i-1] + out[i-2];
        }
    }

  Rcout << "done." << std::endl;
  return out;
}
```

```{r}
sourceCpp('~/Stats506/FibCPP2.cpp')
FibCpp2(10)
```

[Here]() is another simple example and an exercise. 

