---
title: "Dplyr vs Data.Table"
author: "Statistics 506, Fall 2017"
date: ""
output: pdf_document
fontsize: 12pt
geometry: margin=1in
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
library(tidyverse)
library(data.table)
nyc14 = fread('https://github.com/arunsrinivasan/flights/wiki/NYCflights14/flights14.csv')
```

## Instructions
Consider the 2014 flights data used for the `data.table` examples. For each
code snippet first provide a single-sentence summary of the value(s) being computed.
Make your answer as concise and descriptive as possible. 

## Questions 
1.  Consider the following `data.table` expression:

```{r}
tab1 =
  nyc14[ , .(n = .N), by=.(origin, dest)] %>%
    .[, .(origin, n = n, pct = n / sum(n)), by=dest] %>%
    .[pct > .75] %>%
    .[order(-pct, dest)]
```

 a. Provide a one-sentence summary of what is being computed.
\vspace{.2in}    
 b. Provide a translation using `dplyr` syntax.
\vspace{.75in}    

2. Consider the `dplyr` code snippet below.  

```{r}
tab2 =
  nyc14 %>%
    group_by(origin, dest, carrier) %>%
    summarize(n = n()) %>%
    filter(n >= 80) %>%
    group_by(origin, carrier) %>%
    summarize(n = n()) %>%
    arrange(origin,-n)
```
  
 a. Provide a one-sentence summary of what is being computed.
\vspace{.2in}    
 b. Provide a translation using `data.table` syntax.

\pagebreak

  3.  Consider the R code snippet below.
  
```{r}
nyc14_df = as.data.frame(nyc14)
tab3 =
 with(
  with(nyc14_df,nyc14_df[grepl('HOU',dest),]),
      {
       keys = paste(carrier,month,sep=':')
       u = unique(keys)
       n = sapply(u, function(key) sum(key==keys))

       tmp = strsplit(u,':')
       carrier = sapply(tmp,function(x) x[1])
       month  = sapply(tmp,function(x) x[2])
       
       cr  = unique(carrier)
       ind = sapply(cr, function(x) grep(x,carrier))
       data.frame(carrier,month,n)[ind,]
      }
 )
```

 a. Provide a one-sentence summary of what is being computed.
\vspace{.25in}    
 b. Provide a translation using `data.table` syntax.
\vspace{1.25in}    
 c. Provide a translation using `dplyr` syntax.


