---
title: "Introduction to SQL"
author: "Statistics 506, Fall 2017"
date: ""
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## SQL and Relational Databases

A [relational database](https://en.wikipedia.org/wiki/Relational_database)
is a set of rectangular data frames called _tables_
linked by _keys_ relating one table to another.  Software implementations of
such data structures are known as _relational database management systems_ (RDBMS).
Most RDBMS use _structured query language_ ([SQL](https://en.wikipedia.org/wiki/SQL)) 
to modify or search the relational database.

SQL provides a syntax for interfacing with relational data. It is largely a 
declarative language in that we use SQL to specify _what_ we wish to accomplish,
leaving the _how_ to the RDBMS.  While there are standards for SQL implementations
put out by the International Organization for Standardization (ISO) and
the American National Standards Institute (ANSI), there are several open source
and commercial implementations that each have unique features. 

I will try to focus on the commonalities, but will be using an 
[SQLlite](https://en.wikipedia.org/wiki/SQLite) engine in R for providing 
examples. One unique feature of SQLite is that it does not follow the client-server model.
In this model, a physical computer storing the data and executing queries within the RDMBS, 
the _server_, is separate from the machine requesting the queries known as the _client_.
The client-server model is popular in business, health care, and other domains as it allows security
and monitoring of how the data is queried.  It is also popular for many large open
data projects (i.e [ensemble](https://www.ensembl.org/info/data/mysql.html)) where
it is beneficial for data to be centrally maintained and frequently accessed on the fly.

In order to write SQL queries within R I will be using the package
[dbplyr](https://cran.r-project.org/web/packages/dbplyr/vignettes/dbplyr.html)
which provides an SQL back end to `dplyr` functions, but also
allows SQL queries to be parsed as strings.

### Lahman Examples
For our examples today, we will use the `Lahman` package from R which
contains historical baseball data from 1876-2016.

```{r Lahman, warning=FALSE, message=FALSE}
# Packages
library(tidyverse)
library(dbplyr)
library(Lahman)

# Create a local SQLlite database of the Lahman data
lahman <- lahman_sqlite()

# Query the batting table
lahman %>% tbl("BATTING")

# Query 'LahmanData' with meta data
print(lahman %>% tbl('LahmanData') %>% collect(), n = 25)
```

## Working with tables

### Search, subset, and limiting clauses

The basic structure of a SQL query contains
a `SELECT` statement indicating which columns are desired
and a `FROM` statement explaining where to find them. 
You can indicated a subset of rows using a `WHERE` clause.

```sql
CREATE TABLE my_table AS 
SELECT col1, col2, col3 
  FROM some_existing_table
  WHERE col1 IN ('A', 'B', 'C')
```

In the model above, `CREATE TABLE` will only work with a client-server model
if you have permissions to write to memory. Use a wild card `*` to select
all columns in a table:

```{r}
lahman %>% 
  tbl(sql('
      SELECT *
        FROM master
        LIMIT 5
      ')) %>% collect()
```

Above we use a `LIMIT` clause to grab just the first 5 rows of our table. This
is good practice when developing new queries as it prevents large wait times
only to discover a bug.

### Limiting Clauses

Here we will explore the limiting clause `WHERE` in greater detail. Most of what 
we discuss in this section will also work with `HAVING` which operates
on summary statistics rather than individual rows.

#### Boolean Logic

We can use Boolean comparisons in a `WHERE` clause as shown in the example
below. We find all player-seasons since 2000 in which the player was 
credited with an RBI 100 or more times.

Here is our query:

```sql
SELECT playerID, yearID, RBI 
  FROM batting 
 WHERE RBI >= 100 AND yearID >= 2000
```

And here it is in `dbplyr`:

```{r rb100}
## Get all 100+ RBI seasons since 2000
rbi100 =
  lahman %>% 
  tbl(sql('
      SELECT playerID, yearID, RBI 
        FROM batting 
        WHERE RBI >= 100 AND yearID >= 2000
      '))
print(rbi100)
```

Note that the code above creates and test the query, but does not execute it over
the entire table. To do that, we use `collect()`.

```{r rbi100_collect}
## This is dblyr for peforming the query
rbi100 %>% collect()
```

#### IN

To select on a column by testing against a set of fixed values use `IN`.

```sql
SELECT nameGiven as Given, nameLast Last, birthYear AS born --Note equivalent ways to rename columns
  FROM master
 WHERE nameLast IN ("Alou", "Griffey")
```

```{r IN}
## Find all players with last name Alou or Griffey
lahman %>% 
  tbl(sql('
      SELECT nameGiven as Given, nameLast Last, birthYear AS born
        FROM master
        WHERE nameLast IN ("Alou", "Griffey")
      ')) %>% collect()
```
 
#### LIKE

Use a `LIKE` statement with a `WHERE` clause to get partial string matching.
You can use `%` to match any sub-string.

```sql
SELECT nameGiven as Given, nameLast Last, birthYear AS born
  FROM master
 WHERE nameLast LIKE "%riff%" 
```

```{r LIKE}
## Find all players with last name containing 'riff' substring
lahman %>% 
  tbl(sql('
      SELECT nameGiven as Given, nameLast Last, birthYear AS born
        FROM master
        WHERE nameLast LIKE "%riff%"
      ')) %>% collect()
```

Most SQL implementations also have a `REGEXP` or `REGEXLIKE` function that works with
regular expressions, but SQLite requires a [user defined](https://stackoverflow.com/questions/5071601/how-do-i-use-regex-in-a-sqlite-query) `regex()` for its use so
we skip it here.

### Combining limiting clauses

Limiting `WHERE` clauses can be combined using `AND` and `OR`. Statements
can be negated using `NOT`.

```sql
SELECT nameGiven as Given, nameLast Last, birthYear AS born, birthCountry as Place
  FROM master
 WHERE birthCountry == "P.R." AND birthYear LIKE "199%"
```

```{r AND}
## Find all players born in Puerto Rico during the 1990's
lahman %>% 
  tbl(sql('
      SELECT nameGiven as Given, nameLast Last, birthYear AS born, birthCountry as Place
        FROM master
        WHERE birthCountry == "P.R." AND birthYear LIKE "199%"
      ')) %>% collect()
```

```{r OR}
## Find all players born in Alaska (AK) or Hawaii (HI)
lahman %>% 
  tbl(sql('
      SELECT nameGiven as Given, nameLast Last, birthYear AS born, birthCountry as Place
        FROM master
        WHERE birthState == "HI" OR birthState == "AK"
      ')) %>% collect()
```

### ORDER BY

Use an `ORDER BY` clause with a comma separated list of columns to arrange
the table.

```sql
SELECT playerID, yearID, RBI 
  FROM batting 
 WHERE RBI >= 100 AND yearID >= 2010
 ORDER BY yearID, RBI
```

```{r rb100_order}
## Get all 100+ RBI seasons since 2010, ordered 
rbi100 =
  lahman %>% 
  tbl(sql('
      SELECT playerID, yearID, RBI 
        FROM batting 
        WHERE RBI >= 100 AND yearID >= 2010
        ORDER BY yearID, -RBI
      '))
rbi100 %>% collect()
```

## Aggregations, Group By

We can perform aggregations such as sums, means, and counts similar to `summarize()`
from `dplyr`.

We can use a `GROUP BY` clause for the split-apply-combine pattern.

Here we find the players with the most total RBI since 2010.

```sql
SELECT playerID, sum(RBI) as RBI_total
  FROM batting 
  WHERE yearID >= 2010
  GROUP BY playerID
  ORDER BY -RBI_total
```

```{r rb100_total}
## Count total RBIs since 2010 by player
lahman %>% 
  tbl(sql('
      SELECT playerID, sum(RBI) as RBI_total
        FROM batting 
        WHERE yearID >= 2010
        GROUP BY playerID
        ORDER BY -RBI_total
      ')) %>% collect()
```

### Having

The operator defining a limiting clause on an aggregate variable is `HAVING`. It
is essentially like `WHERE` except for operating on summary statistics rather than
individual rows.

In the query below, observe that the `HAVING` clause comes after the `GROUP BY`

but before the `ORDER BY`.
```sql
SELECT playerID, sum(RBI) as RBI_total
  FROM batting 
 WHERE yearID >= 2010
 GROUP BY playerID
   HAVING RBI_total >= 500
 ORDER BY -RBI_total
```

```{r rbi500}
## Players with 500+ RBIs since 2010
lahman %>% 
  tbl(sql('
      SELECT playerID, sum(RBI) as RBI_total
        FROM batting 
        WHERE yearID >= 2010
        GROUP BY playerID
        HAVING RBI_total >= 500
        ORDER BY -RBI_total
      ')) %>% collect()
```
 
## Joins

So far we have discussed working with single tables only.  The SQL term for
merging data from two or more tables is a 'join'.  All joins are based
on the idea of equating rows that match on one or more variables.  Below we will discuss the
following types of joins:

  + Inner Joins - produce tables containing rows for which matches are found in both tables,
  + Left Joins - produce tables containing rows for (at least) each row on the left table with additional
  columns from the right table,
  + Right Joins - produce tables like left joins but reversing the role of the right and left tables
  + Full or Outer Joins - produce tables with all rows appearing in either table.
  + Anti Joins - produce tables with all unmatched rows from both tables.

### Inner Join

What if we wanted to supplement our earlier table showing players with
500+ RBI since 2000 with information about those players? We could use
an inner join of our RBI table with the 'Master' table to accomplish this.

```sql
SELECT m.nameFirst First, m.nameLast Last, sum(RBI) as RBI_TOTAL
  FROM batting b
  INNER JOIN master m ON b.playerID = m.playerID 
  WHERE yearID >= 2010
  GROUP BY b.playerID
  HAVING RBI_total >= 500
  ORDER BY -RBI_total
```

```{r inner_join}
lahman %>% tbl(sql(
'
SELECT m.nameFirst First, m.nameLast Last, sum(RBI) as RBI_TOTAL
  FROM batting b
  INNER JOIN master m ON b.playerID = m.playerID 
  WHERE yearID >= 2010
  GROUP BY b.playerID
  HAVING RBI_total >= 500
  ORDER BY -RBI_total
'
))
```

In this example, an inner join suffices because each `playerID` in the `batting`
table is linked to a `playerID` in the `master` table. 

### Left & Right (Outer) Joins

In a left join -- sometimes called a left outer join -- we add columns
from the right table to the left table when matching rows are found.  Rows
from the left table with no matches from the right table are retained with
columns from the right table filled in as `NULL` (i.e. `NA`).  When there
are multiple matches of a row from the left table to rows in the right table, 
these each become a row in the new table.

A right join is equivalent to a left join with the exception that the roles
between right and left are reversed. My personal preferences is to always structure
queries as left joins for simplicity.  

Left joins are particularly useful when the information in the right table is only 
applicable to a subset of the rows from the left table. As an example,
suppose we would like to know which US colleges and universities have produced
the most "Rookie of the Year Awards" given to the best debuting player(s) each season.

To get started, we first test a query to find the last college attended.
```sql
-- Last college attended
SELECT *
FROM CollegePlaying
GROUP BY playerID
HAVING yearID == max(YearID)
```

```{r last_col}
# Query to find last college atttended
lahman %>% tbl(sql(
'
SELECT *
FROM CollegePlaying
GROUP BY playerID
HAVING yearID == max(YearID)
'
))
```

Now, we find all distinct awards in the `AwardPlayers` table.
```{r awards}
# Distinct Player Awards
lahman %>% tbl(sql(
'
SELECT distinct(awardID)
FROM AwardsPlayers
'
)) %>% collect() %>% print(n=30)
```


Next we test a query for finding all Rookie of the Year Awards.
```sql
SELECT *
FROM AwardsPlayers
WHERE awardID LIKE "Rookie%"
```

```{r roy}
# Query to find Rookie of the Year Awards
lahman %>% tbl(sql(
'
SELECT *
FROM AwardsPlayers
WHERE awardID LIKE "Rookie%"
'
)) 
```

Finally, we use a _left join_ of the tables for ROY awards and last college
attended to match winners to their schools.  We need a left join as many 
of the winners may never have played collegiate baseball.

```sql
SELECT roy.playerID playerID, roy.yearID year, lgID league, schoolID
  FROM AwardsPlayers roy
  LEFT JOIN 
    (SELECT *  --Final College Attended
      FROM CollegePlaying
      GROUP BY playerID
      HAVING yearID == max(YearID)
    ) c
 ON c.playerID = roy.playerID
 WHERE awardID LIKE "Rookie%"
```

```{r last_college_roy}
# Query to find last college for ROY
lahman %>% tbl(sql(
'
SELECT roy.playerID playerID, roy.yearID year, lgID league, schoolID
  FROM AwardsPlayers roy
  LEFT JOIN 
    (SELECT *  --Last College Attended
      FROM CollegePlaying
      GROUP BY playerID
      HAVING yearID == max(YearID)
    ) c
 ON c.playerID = roy.playerID
 WHERE awardID LIKE "Rookie%"
'
))
```

To complete the example, we modify the query to display which schools
have produced the most ROY awards in total.

```sql
SELECT schoolID, count(c.playerID) as ROY_awards
  FROM AwardsPlayers roy
  LEFT JOIN 
    (SELECT *  --Last College Attended
      FROM CollegePlaying
      GROUP BY playerID
      HAVING yearID == max(YearID)
    ) c
 ON c.playerID = roy.playerID
 WHERE awardID LIKE "Rookie%"
   AND schoolID IS NOT NULL
 GROUP BY schoolID
 HAVING ROY_awards > 1
 ORDER BY -ROY_awards
```

```{r roy_by_school_top}
# Which schools have produced the most ROY?
lahman %>% tbl(sql(
'
SELECT schoolID, count(c.playerID) as ROY_awards
  FROM AwardsPlayers roy
  LEFT JOIN 
    (SELECT *  --Last College Attended
      FROM CollegePlaying
      GROUP BY playerID
      HAVING yearID == max(yearID)
    ) c
 ON c.playerID = roy.playerID
 WHERE awardID LIKE "Rookie%"
   AND schoolID IS NOT NULL
 GROUP BY schoolID
 HAVING ROY_awards > 1
 ORDER BY -ROY_awards
'
)) %>% collect()
```

## Using `show_query()`

If you are new to SQL but comfortable with `dplyr` you may find `show_query()` useful
for understanding how dplyr statements are translated to SQL.

```{r}
# Players with more than 5 years at a single school.
lahman %>% 
  tbl('CollegePlaying') %>% 
  filter(yearID > 1900) %>%
  group_by(playerID, schoolID) %>%
  summarize(n_year = n()) %>%
  filter(n_year>5) %>%
  show_query()
```


## Resources 

+ [SQL Tutorials](https://www.w3schools.com/sql/)

+ [SQL Joins using data.table](https://rstudio-pubs-static.s3.amazonaws.com/52230_5ae0d25125b544caab32f75f0360e775.html)

+ [SQL in 10 Easy Steps](https://blog.jooq.org/2016/03/17/10-easy-steps-to-a-complete-understanding-of-sql/)

+ [Databases using dplyr](http://db.rstudio.com/dplyr/)