---
title: "String Manipulation"
author: "Statistics 506, Fall 2017"
date: ""
output: 
  html_document:
    theme: journal
    highlight: pygments
    css: styles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Working with Strings

In R, you create strings of type `character` using either single or double quotes.
There is no difference (in R) between the two.

```{r}
string1 = "This is a string."
string2 = 'This is a string.'
all.equal(string1, string2)
typeof(string1)
```

This is not the case in all languages. For instance, consider the
following example from `bash`.

```
#!/bin/bash
## A short script to illustrate single vs double quotes

# Specify filename and extension
FILE=my_file
EXT=.txt

# Double quotes allow for expansion
echo "Double quotes: $FILE$EXT"

# Single quotes create a literal string
echo 'Single quotes: $FILE$EXT'
```

```
Double Quotes: my_file.txt
Single Quotes: $FILE$EXT
```

Returning to R, you can mix single and double quotes when you want to include
one or the other within your string.

```{r}
string_single = "These are sometimes called 'scare quotes'."
print(string_single)

string_double = 'Quoth the Raven "Nevermore."'
print(string_double)
cat(string_double,'\n')
```

You can also include quotes within a string by escaping them:
```{r}
string_double = "Quoth the Raven \"Nevermore.\""
print(string_double)
cat(string_double,'\n')
```

Observe the difference between `print()` and `cat()` in terms of how the escaped
characters are handled. Be aware also that because backslash plays this special role
as an escape character, it itself needs to be escaped:

```{r}
backslash = "This is a backslash '\\', this is not '\ '."
writeLines(backslash)
```

### String operations

The table below collects some common string operations from base R and their
parallels in the `stringr` package.  I say *parallels* and not *equivalents* because
they do not always behave in the same way.  If you are getting started in R, I suggest
you use `stringr` for these operations.  You should however be aware of the common
base functions as you may encounter them in code written by others.

operation | base | stringr 
---- | ---- | ----
join | `paste` | `str_c`
subset | `substr` | `str_sub`
split | `strsplit` | `str_split`
search | `grep`, `grepl` | `str_locate`,`str_detect`

#### concatenating strings
The functions `paste` and `stringr::str_c` are both used to join strings together.

Observe the difference between the `sep` and `collapse` arguments in `paste`.
```{r}
length(LETTERS)
paste(LETTERS,collapse="")
paste(1:26,LETTERS,sep=': ')
paste(1:26,LETTERS,sep=': ',collapse='\n ')
```

Below we see that `str_c` behaves similarly.
```{r}
library(stringr)
all.equal(str_c(LETTERS,collapse=""), paste(LETTERS, collapse=""))
all.equal(str_c(1:26,LETTERS,sep=': '), paste(1:26, LETTERS,sep=': '))
all.equal(str_c(1:26,LETTERS,sep=': ', collapse='\n '), paste(1:26, LETTERS,sep=': ', collapse='\n '))
```

However, these functions differ in the treatment of missing values  (`NA`).

```{r}
paste(1:3,c(1,NA,3),sep=':', collapse=', ')
str_c(1:3,c(1,NA,3),sep=':', collapse=', ')
str_c(1:3, str_replace_na(c(1,NA,3)), collapse=', ')
```

#### length

Recall that `length` returns the length of a vector. To get the length of a string
use `nchar` or `str_length`.

```{r}
length(paste(LETTERS,collapse=""))
nchar(paste(LETTERS,collapse=""))
str_length(paste(LETTERS,collapse=""))
```

#### substrings

The following functions extract sub-strings at given positions.
```{r}
substr('Strings',3,7)
str_sub('Strings',1,6)
```

The function `stringr::str_sub` supports negative indexing.
```{r}
sprintf('base: %s, stringr: %s', substr('Strings',-5,-1), str_sub('Strings',-5,-1))
```

#### finding matches
The example below uses the vector `fruit` from the `stringr` package.

The base function `grep` returns the indices of all strings within a vector that
contain the requested pattern.
The `grepl` function behaves in the same way but returns a logical vector of the
same length as the input `x`.

```{r}
head(fruit)
grep('fruit', fruit)
which(grepl('fruit', fruit))
head(grepl('fruit', fruit))

```

These functions are vectorized over the input but not the pattern.

```{r}
grep(c('fruit', 'berry'), fruit)
sapply(c('fruit', 'berry'), grep, x=fruit)
```

The `match` function is vectorized over the input, but returns only the first match 
and requires exact matching.

```{r}
match('berry',fruit)
match(c('apple', 'pear'), c(fruit,fruit))
```

The corresponding `stringr` functions are vectorized over both pattern and input,
but the vectorization uses broadcasting so be careful.
Pay attention to the order that the string and pattern are supplied in.

```{r}
ind_fruit = which(str_detect(fruit, 'fruit'))
ind_berry = which(str_detect(fruit, 'berry'))

ind_either = which(str_detect(fruit, c('fruit','berry')))
setdiff(union(ind_fruit, ind_berry), ind_either)

ind_odd = seq(1, length(fruit), 2)
ind_even = seq(2, length(fruit), 2)

odd_fruit = ind_odd[str_detect(fruit[ind_odd], 'fruit')]
even_berry = ind_even[str_detect(fruit[ind_even], 'berry')]
setdiff(union(odd_fruit, even_berry), ind_either)
```

The vectorization in this case doesn't help us to avoid the `lapply` pattern we
used with `grep`.

```{r}
sapply(c('fruit', 'berry'), function(x) which(str_detect(fruit,x)))
```

However, `str_locate` is vectorized using an "OR" operator.

```{r}
ind_fruit = str_locate(fruit, 'fruit')
ind_berry = str_locate(fruit, 'berry')

ind_either = str_locate(fruit, c('fruit','berry'))
setdiff(union(ind_fruit, ind_berry), ind_either)
```

#### splitting strings

The base function `strsplit` can be used to split a string into pieces based on 
a pattern. The example below finds all two-word fruit names from `fruit`.

```{r}
fruit_list = strsplit(fruit,' ')
two_ind = which(sapply(fruit_list, length)==2)
fruit_two = lapply(fruit_list[two_ind], paste, collapse=' ')
unlist(fruit_two)
```

```{r}
all.equal(fruit_list, str_split(fruit, ' '))
```

When there are multiple patterns matching the split point, these `strsplit` and
`str_split` behave differently.

```{r}
string = '1;2;3'
strsplit(string, ';')
str_split(string,';')
str_split(string,';')
```

## Regular Expressions

*Regular expressions* are a way to describe patterns in strings.  There is a common
regexp vocabulary though some details differ between languages.  
The basic idea is illustrated in the examples below using the fruit data.

```{r}
## find all two word fruits
fruit[grep(" ",fruit)]

## find all fruits with an 'a'
fruit[grep("a", fruit)]

## find all fruits starting with 'a'
fruit[grep("^a", fruit)]

## find all fruits ending with 'a'
fruit[grep("a$", fruit)]

## find all fruits starting with a vowel
fruit[grep("^[aeiou]", fruit)]

## find all fruits with two consecutive vowels
fruit[grep("[aeiou]{2}", fruit)]

## find all fruits ending with two consecutive consonants other than r
fruit[grep("[^aeiour]{2}$", fruit)]
```

In the examples above, we return all strings matching a simple pattern. We can 
specify that the pattern be found at the beginning `^a` or end `a$` using *anchors*.
We can provide multiple options for the match within brackets `[]`. We can negate
options within brackets using `^` in a different context. The curly braces ask for 
a specific number (or range `{min, max}`) of matches. 

In the example below we use `.` to match any (single) character. This behaves
much like `?` in Unix file names. We can ask for multiple matches by appending
`*` if we want 0 or more matches and `+` if we want at least 1 match.

```{r}
## find all fruits with two consecutive vowels twice (?)
fruit[grep("[aeiou]{2}.[aeiou]{2}", fruit)]

## find all fruits with two consecutive vowels twice
fruit[grep("[aeiou]{2}.+[aeiou]{2}", fruit)]

## find all fruits with extactly three consectutive consonants in the middle
fruit[grep("[aeiou][^aeiou ]{3}[aeiou]", fruit)]
```

To match an actual period, use the regular expression `\\.`
```{r}
c(fruit, "umich.edu")[grep('\\.', c(fruit, "umich.edu"))]
```

The double backslash is needed because the regular expression itself is passed
as a string and strings also use backslash as an escape character.  This is also
important to remember when building file paths as strings on a Windows computer.


Matched values can be grouped using parentheses `()` and referred back to in the
order they appear using a back reference `\\1`. 

```{r}
## find all fruits with a repeated letter
fruit[grep("(.)\\1", fruit)]

## find all fruits with a repeated letter but exclude double r
fruit[grep("([^r])\\1", fruit)]
```


## Resources

  + "String Manipulation" (Chapter 11) in Matloff's _The Art of R Programming_. 

  + The `stringr` [readme](https://github.com/tidyverse/stringr).
  
  + ["Strings"](http://r4ds.had.co.nz/strings.html) (Chapter 14) from Wickham's _R for Data Science_. 
  
  + An interactive [tutorial](https://regexone.com/lesson/introduction_abcs) on regular expressions.
